### 2014-04-11 patchy-selection-paper.tex
ref
hitting times

@@

First, we compute the second term in equation \eqref{eqn:gestalt_q}, (mean occupation density of a family at $x$),
by which we mean the expected total number of offspring of a single family that ever live at distance $r$ from the patch.
In two dimensions, this depends on where the family began, 
and right-hand side of equation \eqref{eqn:gestalt_q} should be integrated over this starting point.
Here we'll write $x_0$ for the starting point, and $u(x_0,x)$ for the mean occupation density at $x$.
Concretely, suppose that $Z$ is a spatial branching process as above, started with a single individual at $x_0$,
defined now so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the set $A$.
The mean occupation density of $Z$ is then defined to be the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x_0,x)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x_0,x) dx .
\end{align}
Since a family has on average $(1-s_m)^t$ members in generation $t$,
and the marginal distribution of any one surviving individual is approximately Gaussian with variance $t \sigma^2$,
exchanging the expectation and the integral shows that
the density $u(x_0,x)$ is equal to the Green's function for Brownian motion killed at rate $1-s_m$,
\begin{align} \label{eqn:occupation_density_solution}
    u(x_0,x) = \int_0^\infty (1-s_m)^t p_{\sigma^2 t}(x_0,x) dt
\end{align}
where $p_t(x,y) = (2 \pi t)^{-d/2} \exp(-\|x-y\|^2/2t)$ is the transition density for standard Brownian motion.
The integral can be expressed in terms of modified Bessel functions:
using \citet{gradshteyn2007table} (3.471.9 and 8.451.6), the mean occupation density is,
with $\alpha = -\log(1-s_m)$ and $d$ the dimension,
\begin{align}
u(x_0,x) &= \frac{\alpha^{(d-2)/2}}{\sigma^d \pi^{d/2}}\left(\frac{ \sqrt{2 \alpha} \|x-x_0\|}{\sigma} \right)^{(2-d)/2} K_{(2-d)/2}\left( \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right) \\
         &= \frac{\alpha^{(d-2)/2}}{\sigma^d 2 \pi^{(d-1)/2}} \left( \frac{\sqrt{2 \alpha} \|x-x_0\|}{\sigma} \right)^{(1-d)/2} \exp\left(- \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right)+ O(\sigma/\|x-x_0\|) . \label{eqn:occupation_asymptotics}
\end{align}
Since we only have asymptotics for the equilibrium frequency, 
we will only use the latter (asymptotic) form.

@@

Now, we can obtain the first term in equation \eqref{eqn:gestalt_q}, the (outflux of families).
Define $\gamma(r_0)$ to be this outflux; 
i.e.\ the mean number of individuals per unit time that are born carrying the $B$ allele closer than $r_0$ to the original patch
but migrate to a larger distance.
Then for any point $x$ further than $r_0$ from the patch, 
the mean occupation density $q(x)$ is equal to the integral of $\gamma$ over all locations $x_0$ at distance $r_0$ from the patch,
multiplied by the mean occupation density at $x$ for migrant families beginning at at location.
This is a more precise statement of the decomposition of the equilibrium frequency from equation \eqref{eqn:gestalt_q}:
\begin{align}
    q(x) &= \gamma(r_0) \int_{B(r_0)} u(y,x) dy,
\end{align}
so that $\gamma(r_0) = q(x) / \int_{B(r_0)} u(y,x) dy$, for any choice of $x$.
XXX Green function for BM killed on disk XXX
Now, using equations \eqref{eqn:eqfreq} and \eqref{eqn:occupation_asymptotics}, we get that
\begin{align}
    \gamma(r_0) \simeq 
\end{align}

@@

>>>>

%%%%%%% %%%%%%%%%%%
\subsubsection*{Putting it together}

We are approximating the dynamics of the focal allele in the region further away than $r_0$ from the patch
as the sum of independent migrant families, each of whose dynamics are given by a spatial branching process.
Call $B(r_0)$ the region closer than $r_0$ to the patch, and $\delta B(r_0)$ its boundary.
Suppose that there is a constant outflux of migrant families of rate $\gamma(x)$ from each point $x \in \delta B(r_0)$,
e.g.\ with locations and times of appearances as the points of a Poisson process.
Roughly speaking, equations \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate}
differ only by substituting the average number of members of a migrant family in the new patch
for the probability that a family has any members in the new patch;
by the discussion of the previous section, these differ only by a constant factor.

First we will make this idea more formal.
Suppose that $Z$ is one such spatial branching process as above, started with a single individual at $x$,
defined so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the set $A$.
The \emph{mean occupation density} of $Z$ is then defined to be the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x,y)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x,y) dy .
\end{align}
We can think of $u(x,y)$ as the expected total number of offspring of a family beginning at $x$
that ever live at $y$.
By independence of reproduction and spatial motion,
$u(x,y)$ is the sum over all times of the expected number of offspring
multiplied by the probability density that spatial motion takes a lineage from $x$ to $y$,
i.e.\
\begin{align}
    u(x,y) = \int_0^t e^{-\alpha t} p_t(x,y) dt .
\end{align}
Now, \eqref{eqn:gestalt_q} is
\begin{align}
    q(y) &= \int_{\partial B(r_0)} \gamma(x) u(x,y) dx  \\
         &= \int_{\partial B(r_0)} \gamma(x) \int_0^t e^{-\alpha t} p_t(x,y) dt dx .
\end{align}

Now consider the probability that a migrant family from $x$ will ever establish in the new patch,
which we assume has relatively small area $A$ and is located near position $y$.
The probability that this family reaches $A$ by time $t$ is $(1-p_e(t))$
multiplied by the probability that the family reaches the patch, given survival.
If $t$ is large then the previous section argues that $(1-p_e(t)) \approx e^{-\alpha t}/\E[K]$,
and the probability of reaching the patch by $t$ approximately the chance that a Brownian path
will have hit the patch by then.
If $w = A^{1/d}$ is the width of the patch, then

@@

XXX OMIT FROM HERE? XXX

First, we compute the second term in equation \eqref{eqn:gestalt_q}, (mean occupation density of a family at $x$),
by which we mean the expected total number of offspring of a single family that ever live at distance $r$ from the patch.
In two dimensions, this depends on where the family began, 
and right-hand side of equation \eqref{eqn:gestalt_q} should be integrated over this starting point.
Here we'll write $x_0$ for the starting point, and $u(x_0,x)$ for the mean occupation density at $x$.
Concretely, suppose that $Z$ is a spatial branching process as above, started with a single individual at $x_0$,
defined now so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the set $A$.
The mean occupation density of $Z$ is then defined to be the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x_0,x)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x_0,x) dx .
\end{align}
Since a family has on average $(1-s_m)^t$ members in generation $t$,
and the marginal distribution of any one surviving individual is approximately Gaussian with variance $t \sigma^2$,
exchanging the expectation and the integral shows that
the density $u(x_0,x)$ is equal to the Green's function for Brownian motion killed at rate $1-s_m$,
\begin{align} \label{eqn:occupation_density_solution}
    u(x_0,x) = \int_0^\infty (1-s_m)^t p_{\sigma^2 t}(x_0,x) dt
\end{align}
where $p_t(x,y) = (2 \pi t)^{-d/2} \exp(-\|x-y\|^2/2t)$ is the transition density for standard Brownian motion.
The integral can be expressed in terms of modified Bessel functions:
using \citet{gradshteyn2007table} (3.471.9 and 8.451.6), the mean occupation density is,
with $\alpha = -\log(1-s_m)$ and $d$ the dimension,
\begin{align}
u(x_0,x) &= \frac{\alpha^{(d-2)/2}}{\sigma^d \pi^{d/2}}\left(\frac{ \sqrt{2 \alpha} \|x-x_0\|}{\sigma} \right)^{(2-d)/2} K_{(2-d)/2}\left( \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right) \\
         &= 
    \exp\left(- \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right)
    \left\{
    \frac{\alpha^{(d-2)/2}}{\sigma^d 2 \pi^{(d-1)/2}} \left( \frac{\sqrt{2 \alpha} \|x-x_0\|}{\sigma} \right)^{(1-d)/2} 
+ O\left( \left(\frac{\sqrt{2\alpha}\|x-x_0\|}{\sigma}\right)^{-(d+1)/2} \right) 
    \right\}
    . \label{eqn:occupation_asymptotics}
\end{align}
Since we only have asymptotics for the equilibrium frequency, 
we will only use the latter (asymptotic) form.

@@

Now, we can obtain the first term in equation \eqref{eqn:gestalt_q}, the (outflux of families).
Define $\gamma(r_0)$ to be this outflux; 
i.e.\ the mean number of individuals per unit time that are born carrying the $B$ allele closer than $r_0$ to the original patch
but migrate to a larger distance.
Then for any point $x$ further than $r_0$ from the patch, 
the mean occupation density $q(x)$ is equal to the integral of $\gamma$ over all locations $x_0$ at distance $r_0$ from the patch,
multiplied by the mean occupation density at $x$ for migrant families beginning at at location.
This is a more precise statement of the decomposition of the equilibrium frequency from equation \eqref{eqn:gestalt_q}:
\begin{align}
    q(x) &= \gamma(r_0) \int_{B(r_0)} u(y,x) dy,
\end{align}
so that $\gamma(r_0) = q(x) / \int_{B(r_0)} u(y,x) dy$, for any choice of $x$.
Now, using equations \eqref{eqn:eqfreq} and \eqref{eqn:occupation_asymptotics}, 
we can obtain that (see Appendix \ref{apx:lesbegue})
the outflux of families 
\begin{align}
    \gamma(r_0) \simeq 
\end{align}

@@

\subsection{Temporary work: outflux of families}
\label{apx:lesbegue}

We have that
\[
    \gamma(r_0) = \frac{q(y)}{\int_{B(r_0)} u(x,y) dx}
\]
for any $y$.  
Take $y=(r,0)$.
Note that after scaling by $\sqrt{2\alpha}/\sigma$,
\begin{align}
    \frac{ u(x,y) }{ q(y) } &\simeq C' \frac{ \|y-x\|^{-(d-1)/2} e^{-\|y-x\|} }{ \|y\|^{-(d-1)/2} e^{-\|y\|} } \\
        &= C' \| \hat x - x/r \|^{-(d-1)/2} e^{-r (\|\hat x - x/r\| - 1)} .
\end{align}
Now let $x = (r_0 \cos \theta, r_0 \sin \theta)$,
so that the expression in the exponent is to, first order in $1/r$,
\begin{align}
    r (\|\hat x - x/r\| - 1) &= r \left( \sqrt{ \left( 1 - \frac{r_0}{r} \cos \theta \right)^2 + \frac{r_0^2}{r^2} \sin^2 \theta } - 1 \right ) \\
            &= - 2 r_0 \cos \theta + O(1/r) .
\end{align}
This implies that, still with $r$, $r_0$, and $\theta$ defined as above,
\begin{align}
    \frac{u(x,y)}{q(y)} \longrightarrow_{r \to \infty} C' e^{-2 r_0 \cos \theta} ,
\end{align}
and hence, by bounded convergence, then \citet{gradshteyn2007table} (3.339),
\begin{align}
    \int_{B(r_0)} \frac{u(x,y)}{q(y)} dx &\to C' \int_0^{2 \pi} e^{-2 r_0 \cos \theta} d\theta \\
                                         &= 2 \pi C' I_0(-2r_0),
\end{align}
where $I_0$ is a modified Bessel function of order zero of the first kind.
The effect of the scaling is to express $r_0$ in natural units, i.e.\ changing the result to $2 \pi C' I_0( - 2 \sqrt{2\alpha} r_0 / \sigma ) $.

\subsection{Temporary work: hitting probs}

Now consider the probability that a migrant family from $x$ will ever establish in the new patch,
which we assume has relatively small area $A$ and is located near position $y$.
The probability that this family reaches $A$ by time $t$ is $(1-p_e(t))$
multiplied by the probability that the family reaches the patch, given survival.
If $t$ is large then the previous section argues that $(1-p_e(t)) \approx e^{-\alpha t}/\E[K]$,
and the probability of reaching the patch by $t$ approximately the chance that a Brownian path
will have hit the patch by then.
If $w = A^{1/d}$ is the width of the patch, then 
this is the chance that a $d$-dimensional Bessel process begun at $\|y-x\|$ hits $w$ by time $t$.
Let $\tau_w$ be the hitting time of $w$,
so, integrating by parts, and using Borodin \& Salminen (2.2.0.1 and 4.2.0.1)
\begin{align}
    \int_0^\infty (1-p_e(t)) \P^x\{ \tau_w \le t \} dt  
    &\approx \int_0^\infty e^{-\alpha t} \P^x\{ \tau_w \le t \} dt  / \E[K] \\
    &= \frac{1}{\alpha \E[K]} \int_0^\infty e^{-\alpha t} \P^x\{ \tau_w \in dt \} \\
    &= \frac{1}{\alpha \E[K]} \E^x[e^{-\alpha \tau_w}] \\
    &= 
    \frac{1}{\alpha \E[K]} \times
    \begin{cases}
        e^{-(x-w) \sqrt{2\alpha}} \qquad & d=1 \\
        \frac{ K_0(x\sqrt{2\alpha}) }{ K_0(w\sqrt{2\alpha}) } \qquad & d=2 
    \end{cases}
\end{align}
Note that the $d=2$ case
\begin{align}
    \frac{ K_0(x\sqrt{2\alpha}) }{ K_0(w\sqrt{2\alpha}) } = \sqrt{\frac{w}{x}} e^{-(x-w)\sqrt{2\alpha}} ( 1 + O(1/w) ).
\end{align}
